The \emph{FrobeniusThresholds} package for \emph{Macaulay2}

### 2019-06-07 Fsing/FThresholdsPaper/FThresholds.tex
Edits to introduction of paper

This latter result is interesting from a computational perspective, in that it provides lower bounds for log canonical thresholds.  Furthermore, though there is a general purpose implementation of an algorithm for computing log canonical thresholds in the \emph{Dmodules} package by A.~Leykin and H.~Tsai \cite{DmodulesSource}, it happens that the function for computing $F$-pure thresholds contained in the \emph{FrobeniusThresholds} package is typically much faster, especially so in lower characteristic.  Note that the  \emph{MultiplierIdeals}  package  by Z.~Teitler, B.~Snapp and C.~Raicu also computes log canonical thresholds in many special cases, including monomial ideals, hyperplane arrangements, generic determinantal ideals, and certain binomial ideals \cite{MultiplierIdealsPackage, MultiplierIdealsArticle}.

@@

\subsection*{Some background and notation}
It is important to note that much of the functionality implemented in the \emph{FrobeniusThresholds} is not restricted to regular ambient rings; see Section \ref{sec.IsFPT} for a prominent example of this.

@@

Let us outline a common way in which natural numerical invariants in prime characteristic commutative algebra are constructed:  For every natural number $e$, associate to some fixed set of data|often, a collection of polynomials and ideals|an integer describing something of relevance that depends on $e$ (e.g., the dimension of some interesting vector space constructed in terms of your initial data).   Finally, normalize this integer by dividing by some power of $p^e$, and then take the limit as the integer $e$ tends to infinity.  The resulting limit, should it exist, should then encode some interesting information about your initial set of data.

Consider a nonzero polynomial $f$ and a natural number $e$.  If $f$ does not vanish at the origin, then set set $\nu_f^{\idealm}(p^e) := \infty$.  Otherwise, $f \in \idealm$, and we instead define
\[ \nu_f^{\idealm}(p^e) := \max \{ n>0 : f^n \notin \idealm^{[p^e]} \} \]
where $\idealm^{[p^e]}$ denotes the $p^e$-th Frobenius power of $\idealm$, that is, the ideal generated by the $p^e$-th powers of the elements of $\idealm$.    It is not very hard to see that this number is exactly the largest power of $f$ that is part of a free basis for the localization of the ambient polynomial ring at the origin over its subring of $p^e$-th powers.  As noted earlier, this number also has a natural interpretation in terms of $F$-purity of pairs.

Following our outline, we define
\[ \ft^{\idealm}(f) := \lim_{e \to \infty} \frac{ \nu_f^{\idealm}(p^e)}{p^e} \]

@@

In the literature, the $F$-pure threshold $\ft^{\idealm}(f)$ is often denoted $\fpt(f)$, for obvious reasons.  However, in this note, we adopt the former notation to avoid any possible confusion with the \emph{Macaulay2} function {\tt fpt} described in Section \ref{sec.FPT}.  Indeed, the command {\tt fpt(f)} need not always output the number $\ft^{\idealm}(f) = \fpt(f)$, though in the case that it cannot, it will provide a reasonably good estimate of this number.

@@

We conclude this subsection by briefly reviewing some natural generalizations.  Suppose that $I$ and $J$ are ideals.  If $I$ is contained in the radical of $J$, then we set
%
\[ \nu_I^J(p^e) = \max \{ n : I^n \notin J^{[p^e]} \}. \]
Otherwise, we set $\nu_I^J(p^e) = \infty$.
This clearly generalizes the quantity $\nu_f^{\idealm}(p^e)$ considered earlier, and we call
\[ \ft^J(I) := \lim_{e \to \infty} \frac{ \nu_f^{\idealm}(p^e)}{p^e} \]
the \emph{$F$-threshold of $I$ with respect to $J$}.  These numbers are also rational, when they are finite.
The value $\ft^{\idealm}(I)$ is called the \emph{$F$-pure threshold of $I$}, and if $I = (f)$ is principal,  $\ft^{J}(f) := \ft^J(I)$ is called the \emph{$F$-threshold of $f$ with respect to $J$}.
Like $F$-pure thresholds, $F$-thresholds can also be defined in terms of its test ideals.

Finally, we point out that the  \emph{FrobeniusThresholds} package builds heavily upon the \emph{TestIdeals} package for \emph{Macaulay2} \cite{TestIdealsPackage, TestIdealsPaper,M2}, which provides a broad range of functionality for effective computation in prime characteristic commutative algebra.  In fact, both this package and \emph{TestIdeals} are derived from an older package \emph{PosChar}.

@@

There are three main sections in this paper, which describe and illustrate the main functionality implemented in the package \emph{FrobeniusThresholds}.
\begin{itemize}
\item[$\circ$] \autoref{sec.Nu} describes the fundamental function {\tt nu}.  As its name suggests, this function is concerned with computing the integers $\nu_I^J(p^e)$ described above.
\item[$\circ$] \autoref{sec.IsFPT} explains the functionality for determining whether a given number is the $F$-pure threshold, or more generally, an \emph{$F$-jumping exponent}.
\item[$\circ$] \autoref{sec.FPT} describes the features of the central function of the package, {\tt fpt}.
\item[$\circ$]  \autoref{sec.FutureDirections} describes our plans for future improvements to this package.
\end{itemize}
%{\color{red} Move this later}

\subsection*{Acknowledgements.}  First and foremost, the authors enthusiastically thank all authors and contributors to the
\emph{FrobeniusThresholds} package.  {\color{red} (CHECK THESE NAMES)}
Our package coauthors are Juliette Bruce and Daniel Smolkin.  Erin Bela, Zhibek Kadyrsizova, Moty Katzman, Sara Malec, and Marcus Robinson also contributed valuable code to the package.

>>>>

This latter result is interesting from a computational perspective, in that it provides lower bounds for log canonical thresholds.  Furthermore, though there is a general purpose implementation of an algorithm for computing log canonical thresholds in the \emph{Dmodules} package by A.~Leykin and H.~Tsai \cite{DmodulesSource}, it happens that the function for computing $F$-pure thresholds contained in the \emph{FrobeniusThresholds} package is typically much faster, especially so in lower characteristic.  Note that the  \emph{MultiplierIdeals}  package  by Z.~Teitler, B.~Snapp, and C.~Raicu also computes log canonical thresholds in many special cases, including monomial ideals, hyperplane arrangements, generic determinantal ideals, and certain binomial ideals \cite{MultiplierIdealsPackage, MultiplierIdealsArticle}.

@@

\subsection{Some background and notation}
It is important to note that much of the functionality implemented in the \emph{FrobeniusThresholds} is not restricted to regular ambient rings; see Section \ref{sec.IsFPT} for a prominent example of this.

@@

Let us outline a common way in which natural numerical invariants in prime characteristic commutative algebra are constructed:  For every natural number $e$, associate to some fixed set of data|often, a collection of polynomials and ideals|an integer describing something of relevance that depends on $e$ (e.g., the dimension of some interesting vector space constructed in terms of your initial data).   Finally, normalize this integer by dividing by some power of $p^e$, and then take the limit as the integer $e$ tends to infinity.  The resulting limit, if it exists, should then encode some interesting information about your initial set of data.

Consider a nonzero polynomial $f$ and a natural number $e$.  If $f$ does not vanish at the origin, then set set $\nu_f^{\idealm}(p^e) := \infty$.  Otherwise, $f \in \idealm$, and we instead define
\[ \nu_f^{\idealm}(p^e) := \max \{ n>0 : f^n \notin \idealm^{[p^e]} \} \]
where $\idealm^{[p^e]}$ denotes the $p^e$-th Frobenius power of $\idealm$, that is, the ideal generated by the $p^e$-th powers of the elements of $\idealm$.    It is not very hard to see that this number is exactly the largest power of $f$ that is part of a free basis for the localization of the ambient polynomial ring at the origin, over its subring of $p^e$-th powers.  As noted earlier, this number also has a natural interpretation in terms of $F$-purity of pairs.

Following our outline, we define
\[ \ft^{\idealm}(f) := \lim_{e \to \infty} \frac{ \nu_f^{\idealm}(p^e)}{p^e}. \]

@@

In the literature, the $F$-pure threshold $\ft^{\idealm}(f)$ is often denoted $\fpt(f)$, for obvious reasons.  However, in this note, we adopt the former notation to avoid any possible confusion with the \emph{Macaulay2} function {\tt fpt} described in Section \ref{sec.FPT}.  Indeed, the command {\tt fpt(f)} need not always output the number $\ft^{\idealm}(f) = \fpt(f)$, though in the case that it cannot, it will provide a reasonably good estimate for this number.

@@

We conclude this subsection by briefly reviewing some natural generalizations.  Suppose that $I$ and $J$ are ideals.  If $I$ is contained in the radical of $J$, then we set
%
\[ \nu_I^J(p^e) = \max \{ n : I^n \notin J^{[p^e]} \}. \]
Otherwise, we set $\nu_I^J(p^e) = \infty$.
This clearly generalizes the quantity $\nu_f^{\idealm}(p^e)$ considered earlier, and we call
\[ \ft^J(I) := \lim_{e \to \infty} \frac{ \nu_I^J(p^e)}{p^e} \]
the \emph{$F$-threshold of $I$ with respect to $J$}.  These numbers are also rational, when they are finite.
The value $\ft^{\idealm}(I)$ is called the \emph{$F$-pure threshold of $I$}, and if $I = (f)$ is principal,  $\ft^{J}(f) := \ft^J(I)$ is called the \emph{$F$-threshold of $f$ with respect to $J$}.
Like $F$-pure thresholds, $F$-thresholds can also be defined in terms of its test ideals.

Finally, we point out that the  \emph{FrobeniusThresholds} package builds heavily upon the \emph{TestIdeals} package for \emph{Macaulay2} \cite{TestIdealsPackage, TestIdealsPaper,M2}, which provides a broad range of functionality for effective computation in prime characteristic commutative algebra.  In fact, both this package and \emph{TestIdeals} are derived (and have greatly grown) from an older package \emph{PosChar}.

@@

There are three main sections in this paper, which describe and illustrate the main functionality implemented in the package \emph{FrobeniusThresholds}.
\begin{itemize}
\item[$\circ$] \autoref{sec.Nu} describes the fundamental function {\tt nu}.  As its name suggests, this function is concerned with computing the integers $\nu_I^J(p^e)$ mentioned above.
\item[$\circ$] \autoref{sec.IsFPT} explains the functionality for determining whether a given number is the $F$-pure threshold, or more generally, an \emph{$F$-jumping exponent}.
\item[$\circ$] \autoref{sec.FPT} describes the features of the central function of the package, {\tt fpt}.
\item[$\circ$]  \autoref{sec.FutureDirections} outlines our plans for future improvements to this package.
\end{itemize}
%{\color{red} Move this later}

\subsection*{Acknowledgements.}  First and foremost, we enthusiastically thank everyone who helped complete the \emph{FrobeniusThresholds} package. 
Our package coauthors are Juliette Bruce and Daniel Smolkin.  Erin Bela, Zhibek Kadyrsizova, Moty Katzman, Sara Malec, and Marcus Robinson also contributed valuable code to the package.
